Metric Spaces with Expensive Distances

### 2019-04-15 paper/draft_esa.tex
Fix ESA draft Latex compilation

\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
\subjclass{\ccsdesc[500]{Theory of computation~Random search heuristics}}
% \category{dummy}

@@

\section{Introduction}

Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
in a metric space $(\metricspace,\dist)$, 
consider the following standard operations:

\begin{description}
\item[Approximate Nearest Neighbor] Given $\eps>0$ and a point $q\in\metricspace$,
find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
\[\dist(q,p_i)\leq(1+\eps)\dist(q,p_j)\]

\item[Spanner] Given $\eps>0$, compute a weighted graph $G$ with vertices in $\pointset$
such that for any $u,v\in\pointset$, the shortest path distance between $u$ and $v$
is at most $(1+\eps)\dist(u,v)$.
\end{description}

The performance of algorithms for these problems depends on the number of points,
the dimension of the metric space, and the cost $\complexity$ of computing a distance in the metric space.
It is a common assumption to assume $\complexity$ to be a constant; 
There are good reasons for that: the most common case of a metric space
is $\metricspace=\RR^d$ with $d$ some constant, in which case $\complexity$ can be evaluated in $O(d)=O(1)$ time.
Even if $d$ is considered non-constant,
it can always be assumed that $d\leq n$, hence $\complexity$ is at most $O(n)$.
Another typical assumption is that all pairwise distances are part of the input
in which case $\complexity$ is $O(1)$.

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\RR^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered (\cite{cohen2007stability}). For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
\cite{di2016edit}),
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
Another instance is a collection of high-resolution images
endowed with the Wasserstein (or Earth Movers) metric (\cite{rubner2000earth}).

In such situations with expensive distance computations, 
it makes sense to study a different cost model, where only the number of distance computations
is taken into account. For instance, that means that quadratic time operations in terms of $n$
are not counted towards the time complexity, as long as these operations do not query any distance
in $\metricspace$. We also ignore the space complexity in our model.

We will restrict to the case of \emph{doubling spaces}, that is, the doubling dimension
of $\metricspace$ is bounded by a constant. 
In that situation, standard constructions from computational geometry provide partial answers:
Using net-trees~\cite{hm-fast}, we can construct a $\eps$-well-separated pair decomposition (WSPD)~\cite{CK-decomposition} using $O(n\log n)$ distance queries; a WSPD in turn yields
an $\eps$-spanner immediately. Net-trees can also be used to compute approximate nearest neighbors
performing $O(\log n)$ distance computations per query point.
Krauthgamer and Lee \cite{krauthgamer2005black} investigated \textit{black box model},
and proved that ANN search for $\eps < 2/5$ can be done efficiently (i.e., in polylogarithmic time, with polynomial preprocessing
and space) if and only if the dimension is $O(\log \log n)$; their bounds count the number of distance computations.
However, for our relaxed cost model, we pose the question whether simpler constructions achieve
comparable, or even fewer distance computations.

We also propose a slight variant of our model: we assume that we also have access to an (efficient)
$2$-approximation algorithm for the distance queries. Queries to this approximation algorithm
are not counted in the model, hence we can assume that for each pair of points $(u,v)$, we
know a number $A_{u,v}$ with $\dist(u,v)\leq A_{u,v}\leq 2\dist(u,v)$. This induces an approximate ordering
of all distances in the metric space, and it is plausible to assume that such an ordering will simplify
algorithmic tasks on metric spaces, at least in practice.

\myparagraph{Contributions}
%
We propose simple algorithms for spanner construction and approximate nearest neighbor search
and evaluate them theoretically and experimentally in the defined cost model.

Our algorithms are based on the following simple idea: since distance computations are expensive
and should be avoided, we try to obtain maximal information out of the distances that have been computed
so far. 
This information consists of lower and upper bounds for unknown distances, obtained from known distances
by triangle inequality (see Figure~\ref{fig:1st_example}). We remark that updating these bounds involves $\Omega(n^2)$ arithmetic
operations whenever a new distance has been computed, turning the method useless in the standard computational model.

\begin{figure}[h]
\centering
\includegraphics[width=6cm]{\expDistFigPath  intro_example.eps}
\caption{The compute distances are shown as edges in a graph. Note that the exact distance
of $p_1$ and $p_2$ is unknown. The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
constitutes an upper bound on the distance by triangle inequality.
However, we can also infer that $\dist(p_1,p_2)\geq 3$:
otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
would be shorter than the edge $(p_3,p_4)$, again contradicting
triangle inequality.}
\label{fig:1st_example}
\end{figure}

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds (\cite{althofer1993sparse}).
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
spanners on the tested example. Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size.

For approximate nearest neighbor, we devise a simple randomized incremental algorithm and show that
the number of distance queries to find an approximate nearest neighbor is $O(\log n)$ in expectation.
Our proof is based on the well-known observation that the nearest neighbor changes $O(\log n)$ times
in expectation when traversing the sequence of points, combined with a packing argument certifying that
only a constant number of distances needs to be computed in-between two minima.
We also experimentally evaluate our approach and observe that the approach follows 
roughly the theoretical prediction.

\section{Background and Definitions}
%
\myparagraph{Doubling dimension}
A metric space is called \textit{doubling} with \textit{doubling constant} $k$,
if every ball of radius $r$ can be covered by at most $k$ balls of radius $r/2$,
and $k$ is the smallest number having that property.
The \textit{doubling dimension} of a doubling space is defined as $\log k$
(since we usually ignore multiplicative constants, the base of the logarithm is not really important; however,
we always use $\log$ to denote the logarithm with base 2).
It is easy to see that a subspace of a space with doubling dimension $d$ 
is always doubling and has the doubling dimension $O(d)$ (but not necessarily $d$).

We shall need the following lemma, which is just a reformulation of the well-known
packing lemma for doubling spaces (see \cite{smid_2009}, Sect. 2.2).

\begin{lemma}
\label{lem:packing_lemma}
 Let $(\metricspace,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball 
 $B(x,R)$ in $\metricspace$ such that the distance between any two distinct points of $P$ is at least $r$.
 Then 
 \[|P|\leq \left(\frac{4R}{r}\right)^{d}\]

\end{lemma}
\begin{proof}
We can cover $B(x,R)$ with $2^d$ ball of radius $R/2$, each of these balls we can cover with $2^d$
balls of radius $R/4$, etc. Repeating this process $m := \lceil \log \frac{R}{r/2} \rceil$ times, 
we cover
$B(x, R)$ with $2^{md}$ balls of radius at most $r/2$. Since a ball of radius $r/2$ can
contain at most one point from $P$, 
\[|P|\leq 2^{md}= 2^{\lceil \log \frac{R}{r/2} \rceil d}\leq 2^{(1 + \log \frac{R}{r/2})d}=\left(\frac{4R}{r}\right)^{d}.\qedhere\]
\end{proof}

In the following, we will assume throughout that every considered metric space
has a constant doubling dimension.

\myparagraph{Well-separated pair decomposition}
Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
if 
\[
\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
\]
A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
Har-Peled and Mendel \cite{hm-fast} 
introduced the notion of net-trees and
generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
\begin{enumerate}
    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
    \item If $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time. The algorithm uses the net-tree structure.
\end{enumerate}
The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
than net-trees. We mention in passing that cover trees can also be used for building
a spanner (this can be proven with the same methods), and we use cover trees for
building WSPD spanners in one of our implementations.

\section{Algorithms for spanner construction}

\myparagraph{Spanners and known constructions}
%
Let $(\metricspace, \dist)$ be a finite metric space with $n$ points. 
One way to encode the metric space is a complete weighted graph on $\metricspace$,
where the weights correspond to the distances of the points.
A subgraph $G$ of this graph is called a \emph{$(1+\eps)$-spanner} for $(\metricspace,\dist)$ 
if for any pair of points $(u,v)$,
the shortest path distance $d_{uv}$ of $u$ and $v$ in $G$ satisfies $d_{u,v}\leq (1+\eps)\dist(u,v)$.
In other words, the shortest path metric of $G$ is a good approximation of the actual distance for every pair of points.
Clearly, it is a necessary condition that $G$ is connected, hence every spanner must have at least $n-1$ edges.

The \emph{greedy spanner}(\cite{althofer1993sparse}) is a simple algorithm to compute linear-sized spanners:
\begin{algorithmic}
\label{alg:greedy_spanner}
\Function{GreedySpanner}{$P, \eps$}
    \State {$E\gets\emptyset$}
    \State {Sort all pairwise distances of points in $P$}
    \ForAll {pairs $(p_i,p_j)$ in increasing order}
    \State {$d_{ij}\gets$ Shortest path distance in $(P,E)$}
    \If{$d_{ij}>(1+\eps)\dist(p_i,p_j)$}
    \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
    \EndIf
    \EndFor
    \Return {$(P,E)$}
\EndFunction
\end{algorithmic}

The greedy spanner is guaranteed (\cite{althofer1993sparse}) to return a spanner of size $O(n)$
(for constant doubling dimension and fixed $\eps>0$); in experimental study \cite{farshi2009experimental}
it was also shown to return the sparsest graph. However, it clearly
has to compute all $\binom{n}{2}$ pairwise distances in the sorting step;
this means that in our cost model, the greedy spanner has the worst possible
performance.

On the other hand, spanner constructions based on WSPD only compute
$O(n\log n+n \eps^{-d})$ distances 
to construct an $(1+\eps)$-spanner in doubling dimension $d$.
The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
this construction yields a $O(n)$-size spanner using only $O(n\log n)$ distance
computations. However, the algorithm is significantly more involved.

\myparagraph{Blind spanners}
%
We introduce a new framework for constructing spanners
which we call \emph{blind spanners}: the idea is to maintain,
for every pair of points $(p_i,p_j)$,
a lower bound $a_{ij}$ and an upper bound $b_{ij}$ for $\dist(p_i,p_j)$,
initially set to $[0,\infty)$. While there exists some pair for which $\frac{b_{ij}}{a_{ij}}>(1+\eps)$,
we pick one of them, compute its distance and update the lower and upper bounds of
all pairs with respect to the newly acquired information. Here is the pseudocode:

\begin{algorithmic}
\label{alg:blind_spanner}
\Function{BlindSpanner}{$P, \eps$}
    \State {$E \gets \emptyset$}
    \State {$a_{i,j} \gets 0$ for all $1 \leq i,j \leq n$}
    \State {$b_{i,j} \gets \infty$ for all $1 \leq i,j \leq n, i \neq j$}
    \While {$\exists i \neq j : b_{i,j} / a_{i,j} > 1 + \eps$}
    \State {$(i,j) \gets $} \Call{GetNextEdgeToAdd()}{}
    \State {$v \gets \dist(p_i, pj)$}
    \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
    \State \Call{UpdateBounds}{$i, j, v$}
    \EndWhile
\EndFunction
\end{algorithmic}

In this pseudocode we adopt the convention that a positive number divided by 0 is $\infty$
and $\infty$ is larger than any real number,
thus making the predicate in the while loop well-defined.

We give the details of the \textsc{UpdateBounds} procedure next.
Suppose that
$\dist(p_i,p_j)=v\in\RR$ has been computed.
First, we reset $a_{i,j}$ and $b_{j,i}$ to $v$, since the distance
of $p_i$ and $p_j$ is exactly $v$.
To update the upper bound of some entry $b_{k,\ell}$,
we observe that the shortest path from $p_k$ to $p_\ell$ might now
go through the new edge. Hence, we update
\[
    b_{k,\ell}\gets \min_{i,j}\{b_{k,\ell},b_{k,i}+v+b_{j,\ell},b_{k,j}+v+b_{i,\ell}\}
\]
Repeating this for all $k,\ell$ yields the updated upper bounds.
Note that this results in $O(n^2)$ arithmetic operations,
but no distance computation.

For the lower bound, we observe that for any $1\leq k,\ell\leq n$,
\[
    v-b_{k,i}-b_{\ell,j}
\]
is a lower bound for $\dist(p_k,p_\ell)$. Indeed, this follows from
the triangle inequality
%
\[\dist(p_i,p_j)\leq \dist(p_i,p_k)+\dist(p_k,p_\ell)+\dist(p_\ell,p_j)\]
by rearranging terms and plugging in the upper bounds for $\dist(p_i,p_k)$
and $\dist(p_k,p_\ell)$. An analogue bound holds with $i$ and $j$ swapped.

Moreover, the inequalities
\begin{align*}
  a_{j,\ell} - v - b_{k,i}&\leq \dist(p_k,p_\ell)\\
  a_{j,k} - v - b_{j,i}&\leq \dist(p_k,p_\ell)
\end{align*}
hold by triangle inequality, and the same is true with $i$ and $j$ swapped.
This yields $6$ lower bounds for $\dist(p_k,p_\ell)$, and $a_{k,\ell}$
is updated to the maximum of these six lower bounds and its current value.

\myparagraph{Heuristics}
%
The last missing ingredient of our algorithm is the procedure \textsc{GetNextEdgeToAdd},
that is, how to select the next distance to be computed. We propose two natural choices

\begin{description}
\item[\textsc{BlindRandom}] Among all pairs $(i,j)$ where $\frac{b_{i,j}}{a_{i,j}}>(1+\eps)$,
we pick one pair uniformly at random
\item[\textsc{BlindGreedy}] Pick the pair $(i,j)$ which maximizes the ratio $\frac{b_{i,j}}{a_{i,j}}$.
If the maximizing pair is not unique, choose among the maximizing pairs uniformly at random.
\end{description}
%
The idea behind \textsc{BlindGreedy} is that we query an edge for which we know the least,
in that way hoping to gather most additional information about the metric space.
Also, our conventions imply that in \textsc{BlindGreedy} the edges that have $a_{i,j} = 0$ or $b_{i,j} = \infty$
have the highest priority, so the algorithm first ensures that the graph is connected and there are positive
lower bounds for every edge before it will start adding any other edges. Based on this observation,
we also tested variations of the \textsc{BlindRandom} algorithm, where the algorithm
first enforces connectedness and/or lower bounds (i.e., if there are infinite upper bounds,
then the algorithm can only choose one of the corresponding edges, etc).

The next two heuristics assume the existence of a $2$-approximation algorithm for distance
computation. Denoting by $A_{i,j}$ the number satisfying $\dist(p_i,p_j)\leq A_{i,j}\leq 2 d(p_i,p_j)$,
we sort all pairwise distances according to the values $A_{i,j}$.
This yields a roughly sorted sequence of distance, because when $\dist(p_i,p_j)>2\dist(p_k,p_\ell)$,
then $A_{i,j}>A_{k,\ell}$ is guaranteed.
We propose two further heuristics that attempt to make use of this sorted sequence.
\begin{description}
\item [\textsc{BlindQuasiSortedGreedy}] Traverse the pairs in increasing order with respect to $A_{i,j}$.
\item [\textsc{BlindQuasiSortedShaker}] Alternates between pairs with small and large $A_{i,j}$
by traversing in increasing order of $A_{i,j}$ in odd iterations and in decreasing order
in even iterations.
\end{description}

\textsc{BlindQuasiSortedGreedy} tries to mimic the greedy spanner and hence appears as a natural
choice at first sight. However, anticipating the experimental results, the heuristic yields very poor
results. The reason is that no pair acquires useful lower bounds when only short distance are queried
(the greedy spanner does not have this issue because it knows the distance and hence does not need
lower bounds).
Generally speaking, short distances are good for sharp upper bounds, whereas long distances are
useful for lower bounds. This motivates \textsc{BlindQuasiSortedShaker}
which alternates between short and long distances.

\ignore{
The last option that we consider is based on an additional assumption: while
computing $\dist(\cdot, \cdot)$ exactly is costly, we have access to a
cheap approximation algorithm
that gives a $2$-approximation of $\dist$ much faster 
(the output of this approximation algorithm is guaranteed to
lie in $[\dist(p,q), 2 \dist(p,q)]$). In order to simplify notation, we now assume
that all pairwise distances between our points belong to $[1, 2^t]$ for some $t$.
We introduce \textit{buckets} of the form $[2^k, 2^{k+2}]$ for $k = 0 \dots t$;
these intervals are not disjoint, but by calling the approximation algorithm we can assign
each pair $(p_i, p_j)$ to one of the buckets.
Indeed, let $v$ be the output of the approximation algorithm for $\dist(p,q)$,
then we know that the true distance is in $[v/2, v]$. We want to put the pair $(p,q)$ into the bucket
that fully contains this interval, thus we need to find $k$ such that
$2^k \leq v /2$ and $v \leq 2^{k+2}$. These inequalities give $ 2^{k+1} \leq v \leq 2^{k+2}$,
and we take $k = \lfloor \log v \rfloor - 1$.
Afer we found a bucket for each pair $(p_i, p_j)$, 
we have some sort of a weak ordering: we cannot say anything
about two edges that are assigned to the same bucket or to overlapping buckets, but if the buckets
are disjoint, we know which edge is longer. Now we can either imitate the greedy spanner (non-blind) algorithm,
traversing buckets in the order of their left endpoint, and adding edges from them; or we could alternate: in even iterations
we traverse the buckets from left to right, adding edges from each bucket one by one, and in odd iterations
we traverse the buckets from right to left. This alternative makes sense, because we need to have good lower bounds in
order to build a blind spanner, and good lower bounds can be obtained from the triangle inequality, if one of the edges 
in a path connecting $p_i$ and $p_j$ in $G$ is longer than the sum of the lengths of all other edges.
We refer to the former option as \textsc{BlindQuasiSortedGreedy} and to the latter one as \textsc{BlindQuasiSortedShaker}.
}

\section{Experiments on spanners}
We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
experimentally the performance of these heuristics. Clearly, for this metric space, our
cost model is not meaningful since distance comparisons are cheap;
but we picked this environment for controlled experiments.
In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.

\begin{figure}[ht]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
        legend pos=outer north east]

    \addplot+ table [x=n_points, y=edges_quasi_sorted_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind quasi-sorted greedy}

    \addplot+ table [x=n_points, y=edges_quasi_sorted_shaker, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind quasi-sorted shaker}

    \addplot+ table [x=n_points, y=edges_wspd, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{WSPD}

    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind random}

    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind random, lower bound first}

    \addplot+ table [x=n_points, y=edges_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind greedy}

    \addplot+ table [x=n_points, y=edges_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Greedy}

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Number of edges in blind spanners generated by different variants of the blind
    algorithm. Greedy non-blind algorithm and WSPD algorithm are included for comparison. The plot is for normally distributed points
    in dimension 2, $\eps = 0.1$.}
    \label{fig:spanner_sparseness}
\end{figure}

\begin{figure}[ht]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Points,ylabel=\# Edges / \# Points,
        legend pos=outer north east]

    \addplot+ table [x=n_points, y=ratio_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
    \addlegendentry{Blind random, lower bound first}

    \addplot+ table [x=n_points, y=ratio_blind_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
    \addlegendentry{Blind greedy}

    \addplot+ table [x=n_points, y=ratio_greedy, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1_ratio.csv};
    \addlegendentry{Greedy}

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Ratio \# edges / \# points for different variants of spanner 
    algorithms. The plot is for normally distributed points
    in dimension 2, $\eps = 0.1$.}
    \label{fig:spanner_ratio}
\end{figure}

\begin{figure}[ht!]
    \includegraphics[width=0.7\textwidth]{\expDistFigPath blind-greedy-dims.png}
    \caption{Results of blind greedy spanner for different dimensions.}
    \label{fig:blind_greedy_dimensions}
\end{figure}

\begin{figure}[ht!]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Points,ylabel=\# Edges,
        legend pos=outer north east]

    \addplot+ table [x=n_points, y=edges_blind_random, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind random}

    \addplot+ table [x=n_points, y=edges_blind_random_connect_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind random, connect first}

    \addplot+ table [x=n_points, y=edges_blind_random_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind random, lower bound first}

    \addplot+ table [x=n_points, y=edges_blind_random_connect_first_lower_bound_first, col sep=comma] {\expDistDataPath all_methods_dim_2_eps_0.1.txt};
    \addlegendentry{Blind random, connect first, lower bound first}

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Comparison of the four variants of \textsc{BlindRandom} algorithm.}
    \label{fig:blind_rbr_variants}
\end{figure}

\begin{figure}[ht!]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=epsilon,ylabel=\# Edges,
        legend pos=outer north east]

    \addplot+ table [x=Epsilon, y=edges_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
    \addlegendentry{Greedy, dim = 2}

    \addplot+ table [x=Epsilon, y=edges_blind_greedy_dim_2, col sep=comma] {\expDistDataPath eps_dep.csv};
    \addlegendentry{Blind greedy, dim = 2}

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Number of edges in the blind greedy and greedy spanners for different values of $\eps$. Data is for 400 normally distributed points in $\RR^2$ and $\RR^3$.}
    \label{fig:spanner_eps_dependence}
\end{figure}

\begin{figure}[ht]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# Diagrams,
        legend pos=outer north east]

        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
        \addlegendentry{Greedy, $\eps = 0.2$}

        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
        \addlegendentry{Blind greedy, $\eps = 0.2$}

        \addplot+ table [x=n_points, y=ratio_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
        \addlegendentry{Greedy, $\eps = 0.5$}

        \addplot+ table [x=n_points, y=ratio_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
        \addlegendentry{Blind greedy, $\eps = 0.5$}

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Ratio \# edges / \# diagrams for blind greedy spanner 
    algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
    \label{fig:spanner_ratio_mcgill_perturbed}
\end{figure}

\begin{figure}[ht]
    \begin{centering}
\begin{tikzpicture}
    \begin{axis}[xlabel=\# Diagrams,ylabel=\# Edges / \# All distances,
        legend pos=outer north east]

    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Greedy, $\eps = 0.1$}

    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_01_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Blind greedy, $\eps = 0.1$}

    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Greedy, $\eps = 0.2$}

    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_02_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Blind greedy, $\eps = 0.2$}

    \addplot+ table [x=n_points, y=sparseness_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Greedy, $\eps = 0.5$}

    \addplot+ table [x=n_points, y=sparseness_blind_greedy_dim_0_eps_05_q_1, col sep=comma] {\expDistDataPath mcgill_perturbed_blind_greedy.csv};
    \addlegendentry{Blind greedy, $\eps = 0.5$}

\end{axis}
\end{tikzpicture}
\end{centering}
    \caption{Ratio \# edges / \# all distances for blind greedy spanner 
    algorithm. The plot is for 0-dimensional diagrams of perturbed McGill data
    $\eps = 0.1, 0.2, 0.5$ with $W_1$ distance.}
    \label{fig:spanner_sparseness_mcgill_perturbed}
\end{figure}

\begin{table}\centering
    \ra{1.3}
    \begin{tabular}{@{}rrrcrrcrr@{}}\toprule
                          & \multicolumn{2}{c}{$q=1$} & \phantom{abc} & \multicolumn{2}{c}{$q=2$} & \phantom{abc} & \multicolumn{2}{c}{$q=3$}\\
                          \cmidrule{2-3} \cmidrule{5-6} \cmidrule{8-9}
                          & Greedy & Blind greedy && Greedy & Blind greedy && Greedy & Blind greedy \\ \midrule
$\eps = 0.1$\\
\#edges                    & 12755 & 28350         && 32606 & 72933      &&  41875  & 78977 \\
\% of distances computed   & 13.63 & 28.06         && 32.27 & 72.19      &&   41.45  & 78.17 \\
$\eps = 0.2$\\
\#edges                    & 6314 & 19518         && 13560 & 44282       &&  20335  & 52482 \\
\% of distances computed   & 6.25 & 19.32         && 13.42 & 43.83       &&   20.13  & 51.95 \\
$\eps = 0.5$\\
\#edges                    & 2263 & 10414         && 3411 & 20143        &&  4705  & 27154 \\
\% of distances computed   & 2.24 & 10.3          && 3.38 & 19.94        &&   4.66  & 26.88 \\
\bottomrule
\end{tabular}
\caption{Comparison of \textsc{BlindGreedy} and \textsc{Greedy} algorithms for 0-dimensional
diagrams of shapes from McGill dataset with Wasserstein distance $W_q$. The number of diagrams is 450.}
\label{tbl:mcgill_original_blind_greedy_spanner}
\end{table}

We tested the algorithm for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
\begin{enumerate}
    \item In the \textbf{uniform} test set points are sampled uniformly at
        random from the unit cube in $\mathbb{R}^d$.
    \item In the \textbf{normal} test set points are sampled from the standard
        normal distribution in $\mathbb{R}^d$.
    \item In the \textbf{clustered} test set we first sample cluster centers uniformly 
        at random from $[0,10000]^d$, and then we add normally distributed noise around
        each of the centers. The number of clusters is chosen so that each cluster
        contains 50 points.
    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{xi_d})$,
        where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
\end{enumerate}
In all experiments the algorithms that we tested compared in the same way,
so we only present results for the \textbf{uniform} point set in dimension 2.

Figure \ref{fig:spanner_sparseness}
shows the number of edges of the spanner for various variants of 
blind and non-blind spanner constructions.
Note that for all blind spanner variants, the number of computed distances
is equal to the spanner size, while for the non-blind greedy spanner,
this number is always $\binom{n}{2}$ and for WPSD it is lower bounded
by the size of the spanner.
We can see that, even though none of the blind spanners can produce
spanners of the same quality (i.e., sparse) as the standard greedy algorithm,
\textsc{BlindGreedy} and all variants of \textsc{BlindRandom}
perform significantly better than both variants of \textsc{BlindQuasiSorted}.
Figure \ref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
The ideal behavior is demonstrated by the non-blind greedy spanner,
for which this ratio stays practically constant, confirming the linear growth.
None of the blind algorithms seems to have this property, but among them
the blind greedy spanner is the best one. If we assume that the number of edges
is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
linear regression (after taking $\log$).  We give in the table \ref{tbl:regr_coeff_spanner}
the estimated exponents $\alpha$ for \textsc{BlindGreedy} and standard
greedy algorithms. Note that even for the greedy algorithm these estimated
exponents can be significantly larger than 1,
which is explained by the fact that the number of points
on which we computed spanners is not large enough to clearly see
the linear dependence.

\begin{table}[]
\begin{tabular}{|l|l|l|}
\hline
dimension & \textbf{Greedy (non-blind)} & \textbf{Blind greedy} \\ \hline
2         &         1.08                &  1.12                 \\ \hline
3         &         1.24                &  1.41                \\ \hline
4         &         1.42                &  1.77                 \\ \hline
\end{tabular}
\caption{Estimated exponents in the $|E|= C |V|^\alpha$ dependence of the number of edges
on the number of points. The data is for $\eps = 0.1$ and for uniform points.}
\label{tbl:regr_coeff_spanner}
\end{table}

As for different variants of the \textsc{BlindRandom} algorithm,
we note that their performance is almost the same, and the algorithm works
significantly better than \textsc{QuasiSorted} variants, but obviously
worse than the blind greedy variant. There is a consistent, though small,
difference between the variants that do not force lower bounds first and
the other two variants of the \textsc{BlindRandom} (see Figure \ref{fig:blind_rbr_variants}).

>>>>

\keywords{metric spaces, doubling dimension, spanners, approximate nearest neighbor}
\ccsdesc[500]{Theory of computation~Random search heuristics}
\ccsdesc[500]{Theory of computation~Random search heuristics}
%% \category{dummy}

% \category{dummy}

@@

\input{expensive-distances-main.tex}
